The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
John's small comments

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain fraction of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built near the location of the MWA.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially isolated to the centre of the observed disk caused by the Moon, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}. In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

@@

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. These will be referred to as the EoR low and high bands, respectively. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation. This is the residual noise in the visibilities after RFI excision. A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz. The variability over nights is caused by the different celestial observing times and pointing directions, and therefore the difference in apparent brightness of e.g. the Galaxy.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.

@@

Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in the Digital Video Broadcasting (DVB) band 4 (167--174~MHz), which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}. Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

\subsection{Distribution analysis}
Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that observations cover different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI-free (green lines). The EoR high-band curves do not include the night that was contaminated by DTV RFI.

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.

In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes. This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam. The EoR-high distribution has some extra features in its tails that are not smooth. Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes. Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.

@@

\subsection{RFI types}
A variety of RFI events types are observed in the data sets. While there are too many transmitters to show examples for each, it is helpful to understand what kind of events are visible and how they are flagged by the \textsc{aoflagger}. Therefore, a few typical examples are shown.

@@

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DTV signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data are flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation. When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction. Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

@@

The strategy does not perform well on continuous broadband RFI such as DTV, which is occasionally present due to tropospheric ducting or ionospheric activity. To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night. This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}. The current sensitivity of the flagger is enough for calibration and imaging of MWA data. For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels. One way to increase the sensitivity of the flagger, is by executing the flagger on the summed power of multiple baselines. Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.

>>>>

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection and excision, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially restricted to the centre of the Lunar disk, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}. Another study shows that objects in space of about a meter in size, such as satellites and space debris, may also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

@@

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. These will be referred to as the EoR low and high bands, respectively. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz. The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation. This is the residual noise in the visibilities after RFI excision. A few channels in the FM bands show an abnormal standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz. The variability over nights is caused by the different celestial observing times and pointing directions, and therefore the difference in apparent brightness of e.g. the Galaxy.}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations for each observing night. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.

@@

Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in the Digital Video Broadcasting (DVB) band 4 (167--174~MHz), which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. This could be dealt with by accounting for lower SNR and a decreased $uv$ coverage, however putting such a mechanism in place (rather than just deleting affected snapshots entirely) may not be worth the effort. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}. Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

\subsection{Distribution analysis}
Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations studied here, the distributions are shown in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that observations cover different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI-free (green lines). The EoR high-band curves do not include the night that was contaminated by DTV RFI.

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.

In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes. This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam. The EoR-high distribution has some extra features in its tails that are not smooth. Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes. Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.

@@

\subsection{RFI types}
A variety of RFI events are observed in the data sets. While there are too many transmitters to show examples for each, it is helpful to understand what kind of events are visible and how they are flagged by the \textsc{aoflagger}. Therefore, a few typical examples are shown.

@@

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection and excision, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DTV signals are seen $\sim3\%$ of the time, and when present make observing in the 174--195~MHz range impossible. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data are detected and flagged by the \textsc{aoflagger}, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation. When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This requires computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom sufficient to avoid saturation by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction. Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves accuracy.

@@

The current sensitivity of \textsc{aoflagger} is enough for calibration and imaging of MWA data. However, \textsc{aoflagger} does not perform well on continuous broadband RFI such as DTV, which is occasionally present due to tropospheric ducting or ionospheric activity. To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night. This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}. For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels. One way to increase the sensitivity of the flagger, is by executing the flagger on the summed power of multiple baselines. Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.
